The Evolution of Phenotypically Invariant Genetic Networks

### 2018-03-07 kalman_walk_paper.tex
added to disco

@@

\paragraph{Hybridization}
The means of two allopatric populations each of effective size $N_e$ separated for $T$ generations
will be a distance roughly of order $2\sigma_N \sqrt{T/N_e}$ apart along $\optx$.
(Consult figure \ref{fig:conceptual_fig} for a conceptual diagram.)
A population of $F_1$ hybrids has one haploid genome from each,
whose coefficients are averaged,
and so will have mean system coefficients at the midpoint between their means.
The distribution of $F_2$ hybrids will have mean at the average of the two populations,
but will have higher variance.
The variance of $F_2$ hybrids can be shown to increase linearly with the square of the distance between
parental population means
under models of both simple and polygenic traits.
This is suggested by figure \ref{fig:conceptual_fig} and shown in Appendix \ref{apx:seg_cov}.
Concretely, we expect the population of $F_1$s to have variance $\sigma^2_S$ in the selected direction
(the same as within each parental population),
but the population of $F_2$ hybrids will have variance of order $\sigma^2_S + 4 \omega \sigma^2_N T/N_e$,
where $\omega$ is a factor that depends on the genetic basis of the coefficients.
If the optimal set $\allS$ has dimension $q$,
using the polygenic model of appendix \ref{apx:seg_cov}, 
$\omega$ is proportional to the number of degrees of freedom: $\omega = (p-q)/8$.
If each trait is controlled by a single locus, as in figure \ref{fig:conceptual_fig},
the value is similar.

@@

%  $\sigma_N \approx \sigma_S \ll \gamma$, ``isolated population''
Consider instead a much smaller, isolated population
whose genetic variation is primarily constrained by genetic drift,
so that $\sigma_N \approx \sigma_S \ll \gamma$.
Setting $a = (\sigma_N/\gamma)^2$ to be small,
the fitness of $F_2$ hybrids is $\fit_2 \le 1/\sqrt{1 + 4 \omega a T/N_e} \approx 1 - 2 \omega a T/N_e$.
Hybrid fitness seems to drop more slowly in this case in figure \ref{fig:speciation_rates},
but since time is scaled by $N_e$, so speciation may occur \emph{faster} than in a large population.
However, at least in some models \citep{lynch1986phenotypic}, in small populations at mutation-drift equilibrium
the amount of genetic variance ($\sigma_N^2$) is proportional to $N_e$,
which would compensate for this difference, 
perhaps even predicting the rate of decrease of hybrid fitness to be \emph{independent} of population size
for small populations.

@@

In this paper, we use tools from quantitative genetics and control theory
to study the evolution of a mechanistic model of the genotype-phenotype map, 
in which the phenotype is subject to stabilizing selection.
In so doing, we provide an explicit model of
phenogenetic drift \citep{weiss2000phenogenetic} and developmental system drift \citep{true2001developmental}.
In this context, the Kalman decomposition \citep{kalman1963mathematical}
gives an analytical description of all phenotypically equivalent gene networks.
It also implies that nearly all systems are nonidentifiable,
and that, in general, there exist axes of genetic variation unconstrained by natural selection.
The independent movement of separated populations along these axes by genetic drift
can lead to a significant reduction in hybrid viability, and thus precipitate speciation,
at a speed dependent on the effective population size and the amount of genetic variation. 
In this model, at biologically reasonable parameter values,
system drift is a significant -- and possibly rapid -- driver of speciation.
This may be surprising because
hybrid inviability appears as a consequence of recombining different, yet functionally equivalent, mechanisms, and since species are often defined by their unique adaptations or morphologies.

@@

@@

All of these forces -- adaptive shifts, conflict, and network drift --
are plausible drivers of speciation, and may even interact.
Many of our observations carry over to models of directional selection --
for instance, rapid drift along the set of equivalent systems
could be driven by adaptation in a different, pleiotropically coupled system.
Or, reinforcement due to local adaptation might provide a selective pressure
that speeds up system drift.
Furthermore,
while the fitness consequences of incompatibility in any one given network may be small, 
the cumulative impact of system drift across the many different networks an organism relies on may be substantial. 
It remains to be seen how the relative strengths of these forces compare.

>>>>

As our goal here is to sketch out the rough implications of our main results,
keeping the assumptions clearly visible,
we provide often heuristic arguments,
rather than presenting calculations in full multivariate generality.

@@

\paragraph{Hybridization}
The means of two allopatric populations each of effective size $N_e$ separated for $T$ generations
will be a distance roughly of order $2\sigma_N \sqrt{T/N_e}$ apart along $\optx$.
(Consult figure \ref{fig:conceptual_fig} for a conceptual diagram.)
A population of $F_1$ hybrids has one haploid genome from each,
whose coefficients are averaged,
and so will have mean system coefficients at the midpoint between their means.
The distribution of $F_2$ hybrids will have mean at the average of the two populations,
but will have higher variance \citep{barton2001role,rosas2010cryptic,chevin2014niche}.
This ``segregation variance'' of $F_2$ hybrids 
can be shown to increase linearly with the square of the distance between
parental population means
under models of both simple and polygenic traits.
This is suggested by figure \ref{fig:conceptual_fig} and shown by \citet{slatkin1994segregation}
(also see Appendix \ref{apx:seg_cov} for a different derivation).
Concretely, we expect the population of $F_1$s to have variance $\sigma^2_S$ in the selected direction
(the same as within each parental population),
but the population of $F_2$ hybrids will have variance of order $\sigma^2_S + 4 \omega \sigma^2_N T/N_e$,
where $\omega$ is a factor that depends on the genetic basis of the coefficients.
If the optimal set $\allS$ has dimension $q$,
using the polygenic model of appendix \ref{apx:seg_cov}, 
$\omega$ is proportional to the number of degrees of freedom: $\omega = (p-q)/8$.
If each trait is controlled by a single locus, as in figure \ref{fig:conceptual_fig},
the value is similar.

@@

%  $\sigma_N \approx \sigma_S \ll \gamma$, ``isolated population''
Consider instead a much smaller, isolated population
whose genetic variation is primarily constrained by genetic drift,
so that $\sigma_N \approx \sigma_S \ll \gamma$.
Setting $a = (\sigma_N/\gamma)^2$ to be small,
the fitness of $F_2$ hybrids is $\fit_2 \le 1/\sqrt{1 + 4 \omega a T/N_e} \approx 1 - 2 \omega a T/N_e$.
Hybrid fitness seems to drop more slowly in this case in figure \ref{fig:speciation_rates},
but since time is scaled by $N_e$, so speciation may occur \emph{faster} than in a large population.
However, at least in some models \citep{lynch1986phenotypic}, in small populations at mutation-drift equilibrium
the amount of genetic variance ($\sigma_N^2$) is proportional to $N_e$,
which would compensate for this difference, 
perhaps even predicting the rate of decrease of hybrid fitness to be independent of population size
for small populations.

@@

In this paper, we use tools from linear system theory and quantitative genetics
to study the evolution of a mechanistic model of the genotype-phenotype map, 
in which the phenotype is subject to stabilizing selection.
In so doing, we provide an explicit model of
phenogenetic drift \citep{weiss2000phenogenetic} and developmental system drift \citep{true2001developmental}.
In this context, the Kalman decomposition \citep{kalman1963mathematical}
gives an analytical description of all phenotypically equivalent gene networks.
It also implies that nearly all systems are nonidentifiable,
and that, in general, there exist axes of genetic variation unconstrained by natural selection.
The independent movement of separated populations along these axes by genetic drift
can lead to a significant reduction in hybrid viability, and thus precipitate speciation,
at a speed dependent on the effective population size and the amount of genetic variation. 
In this model, at biologically reasonable parameter values,
system drift is a significant -- and possibly rapid -- driver of speciation.
This may be surprising because
hybrid inviability appears as a consequence of recombining different, yet functionally equivalent, mechanisms, and since species are often defined by their unique adaptations or morphologies.

@@

Substantial analytical work has made similar observations
using Fisher's geometric model, 
including Haldane's rule and the $F_2$ ``speciation clock'' 
\citep{barton2001role,fraisse2016genetics,simon2017coadapted}.
These calculations are similar in spirit, 
but differ from our own in that they primarily assume a monomorphic population.
Indeed, our model may provide an additional mechanistic context 
in which Fisher's geometric model is appropriate,
although the typical degree of pleiotropy
and stability of the $G$-matrix are unknown in practice.
Our observations argue for the importance of including neutral directions in these models,
which is not usually done (but see \citet{chevin2014niche}).

@@

Neutral processes are certainly not the only drivers of speciation.
All of these forces -- adaptive shifts, conflict, and network drift --
are plausible drivers of speciation, and may even interact.
Many of our observations carry over to models of directional selection --
for instance, rapid drift along the set of equivalent systems
could be driven by adaptation in a different, pleiotropically coupled system.
Or, reinforcement due to local adaptation might provide a selective pressure
that speeds up system drift.
Furthermore,
while the fitness consequences of incompatibility in any one given network may be small, 
the cumulative impact of system drift across the many different networks an organism relies on may be substantial. 
It remains to be seen how the relative strengths of these forces compare.
